# Dimensionality Reduction and Exploration of the Merged Slides

```{r, echo=FALSE, include=FALSE}
# load required libraries
up_packages = c("dplyr", "ggplot2", "glmGamPoi", "patchwork", "presto",
                "qs", "RColorBrewer", "Rfast2", "scales", "Seurat", "SeuratData")
lapply(up_packages, require, character.only = TRUE)

# set working directory
mainDir = "/home/andrea/19ADC99/AD_spatial/"
knitr::opts_knit$set(root.dir = mainDir)

# load the merged Seurat object
ST011 = readRDS("/home/andrea/Dump/AD_spatial/ST011.rds")
```


## Dimensionality Reduction

Let's run PCA and UMAP on the merged slides to explore the data structure and relationships between the mapped voxels.

```{r, echo=FALSE}

ST011 = Seurat::RunPCA(ST011, verbose = TRUE)
ST011 = Seurat::FindNeighbors(ST011, dims = 1:30)
ST011 = Seurat::FindClusters(ST011, verbose = TRUE)
ST011 = Seurat::RunUMAP(ST011, dims = 1:30)

```

Let's plot the UMAP embedding of the voxels colored by the cluster assignment.

```{r, fig.cap="UMAP merged slides", fig.height=9,  warning=FALSE, message=FALSE}

# UMAP plot
plot1 = Seurat::DimPlot(ST011, reduction = "umap", group.by = c("ident"), label = TRUE)
plot2 = Seurat::DimPlot(ST011, reduction = "umap", group.by = c("orig.ident"), cols = c("indianred", "brown2", "firebrick", "#BCD2EE", "lightskyblue3", "dodgerblue"), label = FALSE)
patchwork::wrap_plots(plot1, plot2, ncol = 2)

```

Indeed, cell groups 9, 10 and 11 seems enriched in the AD slides. There is also an evident separation between control and AD cells in groups 2 and 5.

Let's now overlay the Groups over the slides.

```{r, fig.cap="Slides VS cell groups", fig.height=9,  warning=FALSE, message=FALSE}

# Visualize Layers
Seurat::SpatialDimPlot(ST011, pt.size.factor = 2000, label = TRUE, label.size = 3, ncol = 3)

```

Let's confirm that cell groups 9, 10 and 11 are indeed enriched in the AD slides by highlighting these cells over the slides.

```{r, fig.cap="Cell Group 9 distribution", fig.height=9,  warning=FALSE, message=FALSE}

# group 9
p1 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[1],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
p2 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[2],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
p3 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[3],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
p4 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[4],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
p5 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[5],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
p6 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[6],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 9),
                            pt.size.factor = 2000) + NoLegend()
patchwork::wrap_plots(p1, p2, p3, p4, p5, p6, ncol = 3)
```


```{r, fig.cap="Cell Group 10 distribution", fig.height=9,  warning=FALSE, message=FALSE}
# group 10
p1 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[1],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
p2 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[2],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
p3 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[3],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
p4 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[4],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
p5 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[5],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
p6 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[6],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 10),
                            pt.size.factor = 2000) + NoLegend()
patchwork::wrap_plots(p1, p2, p3, p4, p5, p6, ncol = 3)

```


```{r, fig.cap="Cell Group 11 distribution", fig.height=9,  warning=FALSE, message=FALSE}

# group 11
p1 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[1],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
p2 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[2],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
p3 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[3],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
p4 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[4],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
p5 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[5],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
p6 = Seurat::SpatialDimPlot(ST011, images = names(ST011@images)[6],
                            cells.highlight = CellsByIdentities(object = ST011, idents = 11),
                            pt.size.factor = 2000) + NoLegend()
patchwork::wrap_plots(p1, p2, p3, p4, p5, p6, ncol = 3)

```




[Wang et al.](https://www.nature.com/articles/s41467-024-49133-z) reports as most Differentially Expressed genes in AD vs control for these slides MTRNR2L1, MTRNR2L6, SERPINA3 and DEPP1.
Let's see the expression of these genes across the slides.

```{r, fig.cap="MTRNR2L1 and MTRNR2L6 expression", fig.height=9,  warning=FALSE, message=FALSE}

# MTRNR2L1 and MTRNR2L6 expression
Seurat::SpatialFeaturePlot(ST011, features = c("MTRNR2L1", "MTRNR2L6"), pt.size.factor = 2000)

```


```{r, fig.cap="DEPP1 and SERPINA3 expression", fig.height=9,  warning=FALSE, message=FALSE}

# DEPP1 and SERPINA3 expression
Seurat::SpatialFeaturePlot(ST011, features = c("DEPP1", "SERPINA3"), pt.size.factor = 2000)

```


## Identification of Spatially Variable Genes across groups

We can also compute the most spatially variable genes across the slides using the `FindMarkers` function in Seurat.
We can choose cell groups 2 and group 11 to see which markers are more differentially expressed across these two groups.

```{r, echo=FALSE, eval=FALSE}
ST011 = Seurat::PrepSCTFindMarkers(ST011)
gc()
```

Let's plot the top 3 markers identified

```{r, fig.cap="Top 3 spatially variable markers", fig.height=9, width=36, warning=FALSE, message=FALSE}
de_markers = Seurat::FindMarkers(ST011, ident.1 = 2, ident.2 = 11)
p1 = Seurat::SpatialFeaturePlot(
    ST011,
    features = rownames(de_markers)[1:3], alpha = c(0.1, 1),
    pt.size.factor = 2000, ncol = 6
)
p1

```





Random graph, go!!!

```{r, fig.cap="Genes heatmap", fig.height=9, fig.width=16, warning=FALSE, message=FALSE}

Seurat::DimHeatmap(allen_MTG_reference)

```