# Transfer cell annotations

```{r, echo=FALSE, include=FALSE}
# load required libraries
up_packages = c("dplyr", "ggplot2", "glmGamPoi", "patchwork", "presto",
                "qs", "RColorBrewer", "Rfast2", "scales", "Seurat", "SeuratData")
lapply(up_packages, require, character.only = TRUE)

# set working directory
mainDir = "/home/andrea/19ADC99/AD_spatial/"
knitr::opts_knit$set(root.dir = mainDir)

# load the merged Seurat object
#ST011 = qs::qread("/home/andrea/Dump/AD_spatial/ST011.qsave", nthreads = 6)
#allen_MTG_reference = qs::qread("/home/andrea/Dump/AD_spatial/allen_MTG_reference.PCA.qsave", nthreads = 6)
```


# Transfer cell annotations from the Allen MTG reference to the AD slides.

First step is to identify and transfer gene anchors between the Allen MTG reference and the AD slides.
We need to add SCT normalization to the Allen MTG reference, and then find the anchors. Since `SCTransform` requires the `counts` slot to be filled, we need to copy the `data` slot to the `counts` slot.  

```{r, echo=TRUE, eval=FALSE}
allen_MTG_reference[["RNA"]]$counts = allen_MTG_reference[["RNA"]]$data
allen_MTG_reference = Seurat::SCTransform(allen_MTG_reference, verbose = TRUE)
```



qs::qsave(allen_MTG_reference, "/home/andrea/Dump/AD_spatial/allen_MTG_reference.PCA.qsave", nthreads=8)

anchors = Seurat::FindTransferAnchors(
    reference = allen_MTG_reference,
    query = ST011,
    normalization.method = "SCT",
    recompute.residuals = FALSE
)

