# Differential gene expression analysis

```{r, echo=FALSE, include=FALSE}
# load required libraries
up_packages = c("dplyr", "ggplot2", "glmGamPoi", "patchwork", "presto",
                "qs", "RColorBrewer", "Rfast2", "scales", "scMiko", "Seurat", "SeuratData")
lapply(up_packages, require, character.only = TRUE)

# set working directory
mainDir = "/home/andrea/19ADC99/AD_spatial/"
knitr::opts_knit$set(root.dir = mainDir)

# load the merged Seurat object
ST011 = qs::qread("/home/andrea/Dump/AD_spatial/ST011.predictions.qsave", nthreads = 6)
```


## Differential gene expression analysis

Let's now focus on identitying markers that can distinguish the different cell types in the dataset. Let's use the function `FindAllMarkers`.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# must switch back to SCT to identify variable markers
Seurat::DefaultAssay(ST011) = "SCT"

# find markers
all_markers = Seurat::FindAllMarkers(
    ST011,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.25
)
```

Let's check the top markers for each identified.

```{r, echo=TRUE}
all_markers %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1)

```

Let's visualize the top 6 in a UMAP plot.

```{r, fig.cap="Top 6 markers", fig.height=9, warning=FALSE, message=FALSE}
Seurat::FeaturePlot(ST011, features = c("MT1F", "MT1G", "MT1E", "WIF1", "ETNPPL", "SLC14A1"))
```

Let's visualize the top 10 markers that differentiate between clusters with a heatmap.

```{r, fig.cap="Top markers heatmap", fig.height=9, warning=FALSE, message=FALSE}
# get the genes
top10_genes = all_markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup()

# plot
p1 = Seurat::DoHeatmap(ST011, features = top10_genes$gene) + NoLegend()
p1
```

We can also check the distribution of the top 10 markers across the different samples and groups.


```{r, fig.cap="Dotplot top markers per sample", fig.height=9,  warning=FALSE, message=FALSE}
Seurat::DotPlot(
    ST011,
    features = unique(top10_genes$gene),
    split.by = "orig.ident",
    cols = c("indianred", "brown2", "firebrick", "#BCD2EE", "lightskyblue3", "dodgerblue")
) + RotatedAxis()
```

```{r, fig.cap="Dotplot top markers per cluster", fig.height=9,  warning=FALSE, message=FALSE}
Seurat::DotPlot(
    ST011,
    features = unique(top10_genes$gene),
    split.by = "seurat_clusters",
    cols = c(brewer.pal(8, "Set2"), brewer.pal(8, "Dark2"))
) + RotatedAxis()
```
